---
title: "GABA-AD: mixed models statistics"
output: html_notebook
chunk_output_type: inline
---

```{r chunk_setup, include = FALSE, eval = TRUE}
knitr::opts_chunk$set(echo = T, message = F, warning = F, comment=NA, autodep = F, eval = T, cache.rebuild = F, cache = T, R.options = list(width = 120), fig.width = 8, fig.align = 'center', dev.args = list(bg = 'transparent'), dev = 'svglite')
```

## Effect of pharmacological actvation of GABAARs
### Outcome variables
- **TEPs**: 5 components (= amplitudes)
  - **mean amplitude** across 25% of points from the window of         interest, which are the most distant from 0
  - **latency** calculated from the median point included in         the amplitude calculation
- **MEPs**: 1 component
  - peak-to-peak amplitude
- **RS-EEG**: 3 extracted variables
  - mean power across the lower beta band = **sigma peak**
  - **Alpha Attenuation Coefficient** (eyes closed/eyes open)
  - **Spectral Exponent** of the 1/f slope
        
All outcome variables are presented as the difference between the value post-medication and the value at baseline

### Fixed effects
- **MEDICATION**: alprazolam x placebo
  - binomial = 2 levels
- **STIMULUS**: CS x TS
  - binomial = 2 levels
- Covariate: **rMT change** 
  - continuous value
- Covariate: **baseline value** of outcome variables
  - continuous value

### Random effects
- **SUBJECT**: to account for unknown interpersonal variability

Following pipeline treats each outcome measure separately, as an independent measure

## Data visualization
Load packages
```{r basic_packages, include = FALSE, eval = TRUE}
library(tidyverse)
library(lme4)
library(car)
library(ggplot2)
library(MASS)
library(dplyr)
library(ggpubr)
library(corrplot) 
library(visreg) 
library(rgl)
library(knitr)
library(scatterplot3d)
library(merTools)
library(readr)
```

Set graphic parameters for visualization
```{r graphic pars}
# color palette - red/blue/purple/green(light + dark)
col_palette = c('cadetblue1', 'turquoise4', 'tomato1', 'red3', 'mediumorchid1', 'darkmagenta', 'olivedrab3', 'olivedrab4')
```

Import data in R-compatible long format
```{r import}
data_all <- read.csv('C:/Users/uzivatel/UCL/O365G-NOCIONS - dsulcova/GABA-AD_results/Data/GABA_med_long.csv')
head(data_all, 10)
```

Choose the outcome variable of interest 
```{r choose data}
# define outcome variable of interest
variable <- c('P60', 'amplitude')

# choose columns to keep
vars_keep <- c('subject', 'medication', 'time', 'stimulus', 'peak', 'TEP', 'rMT')

# make into factors
data <- as_tibble(data_all[vars_keep]) 
data$subject <- factor(data$subject)
data$medication <- factor(data$medication)
data$time <- factor(data$time)
data$stimulus <- factor(data$stimulus)

# choose peak of interest
peak_names <- c('P30', 'N45', 'P60', 'N100', 'P180')
index <- which(peak_names == variable[1])
data <- data[data$peak == index, -5] 

# calculate change post-medication
change <- data[data$time == 1, 5] - data[data$time == 0, 5]
data <- data[data$time == 0, -3] %>%
  rename_at(vars('TEP'), ~ as.character('baseline')) %>%
  add_column(change = deframe(change))
rm(change)
head(data, 10)
```

Get basic descriptive statistics of the data 
```{r desctiptive}
describe <- function(x){
  print(round(c(mean = mean(x, na.rm = TRUE),
            median = median(x, na.rm = TRUE),
            min = min(x, na.rm = TRUE),
            max = max(x, na.rm = TRUE),
            std = sd(x, na.rm = TRUE)), digits = 4))}

# check variable at baseline
cat(paste(variable[1],  ' - baseline: \n'))
describe(data$baseline)
cat('grouped by stimulus: \n')
aggregate(baseline ~ stimulus, data = data, FUN = describe)
cat('grouped by medication: \n')
aggregate(baseline ~ medication, data = data, FUN = describe)
```

Visualize the data, test for normality
```{r first look}
# baseline
dist_bl <- data %>%
  ggplot(aes(x = baseline, fill = stimulus)) +
  geom_histogram(color = 'white', alpha = 0.6, position = 'identity') +
  scale_fill_manual(name = 'stimulus',
                    breaks = c(0, 1),
                    labels = c('80 %rMT', '120 %rMT'),
                    values = col_palette[c(8, 6)]) +
  ggtitle(paste(variable[1], ': ','baseline amplitude distribution')) +
  theme(text = element_text(size = 15),
        panel.background = element_rect(fill = 'white'),
        axis.line = element_line(colour = 'black')) 
dist_bl

# baseline - test of normality
shapiro.test(data$baseline[data$stimulus == 0 & data$medication == 0])
shapiro.test(data$baseline[data$stimulus == 0 & data$medication == 1])
shapiro.test(data$baseline[data$stimulus == 1 & data$medication == 0])
shapiro.test(data$baseline[data$stimulus == 1 & data$medication == 1]) 

# change
dist_change <- data %>%
  ggplot(aes(x = change, fill = medication)) +
  geom_histogram(color = 'white', alpha = 0.6, position = 'identity') +
  scale_fill_manual(name = 'medication',
                    breaks = c(0, 1),
                    labels = c('placebo', 'alprazolam'),
                    values = col_palette[c(2, 4)]) +
  ggtitle(paste(variable[1], ': ','amplitude change distribution')) +
  theme(text = element_text(size = 15),
        panel.background = element_rect(fill = 'white'),
        axis.line = element_line(colour = 'black')) 
dist_change

box_change <- data %>%
  ggplot(aes(x = stimulus, y = change,  fill = medication)) +
  geom_boxplot(alpha = 0.6) +
  scale_fill_manual(name = 'medication',
                    breaks = c(0, 1),
                    labels = c('placebo', 'alprazolam'),
                    values = col_palette[c(2, 4)]) +
  scale_x_discrete(labels = c('0' = '80 %rMT', '1' = '120 %rMT')) +
  ggtitle(paste(variable[1], ': ','amplitude change')) +
  theme(text = element_text(size = 15),
        panel.background = element_rect(fill = 'white'),
        axis.line = element_line(colour = 'black')) 
box_change 

# change - test of normality
data_test = data$change[data$stimulus == 0 & data$medication == 0]
shapiro.test(data_test)
qqnorm(data_test)
qqline(data_test)


data_test = data$change[data$stimulus == 0 & data$medication == 1]
shapiro.test(data_test)
qqnorm(data_test)
qqline(data_test)

data_test = data$change[data$stimulus == 1 & data$medication == 0]
shapiro.test(data_test)
qqnorm(data_test)
qqline(data_test)

data_test = data$change[data$stimulus == 1 & data$medication == 1]
shapiro.test(data_test)
qqnorm(data_test)
qqline(data_test)

# homoscedasticity
leveneTest(change ~ medication, data = data[data$stimulus == 0, ], center = median)
leveneTest(change ~ medication, data = data[data$stimulus == 1, ], center = median)
```


