---
title: 'GABA-AD: mixed models statistics'
output: 
  html_notebook:
    theme: yeti
---

```{css, echo = FALSE}
.codeblock {
  background-color: lightblue;
  border-style: none;
  font-color: darkblue;
}
.output {
  border: 2px lightblue;
}
```

```{r chunk_setup, include = FALSE, eval = TRUE}
knitr::opts_chunk$set(message = F, warning = F, comment = NA, 
                      cache.rebuild = F, cache = T, R.options = list(width = 120), 
                      fig.width = 8, fig.align = 'center',
                      class.source = 'codeblock', class.output = 'output')
```

## Effect of pharmacological actvation on outcome variables
### Outcome variables
Each type of variable will be tested as output to 
- TEPs: 5 components (peaks P30, N45, P60, N100, P180) --> mean amplitude 
- MEPs: peak-to-peak amplitude 
- RS-EEG: 3 related variables

### Factors
- MEDICATION: alprazolam X placebo
- TIME: baseline X post medication
- STIMULUS: CS X TS

### Group by
- Participant --> individual (inter-personal) variability --> participant RE
- Session --> intra-personal variability
- Stimulus --> only for TEPs, to cover for the fact that all TEP components are related

## Statistical pipeline
Load packages
```{r basic_packages, echo = TRUE, include = FALSE}
library(tidyverse)
library(lme4)
library(car)
library(ggplot2)
library(MASS)
library(dplyr)
library(ggpubr)
library(corrplot) 
library(visreg) 
library(rgl)
library(knitr)
library(scatterplot3d)
library(merTools)
library(readr)
```


Import data in R-compatible long format
```{r load_data, class.output = 'success'}
filename = file.choose()
data = read.csv(filename)
head(data, 20)
```

 
```{r}


```