---
title: "GABA-AD: mixed models statistics"
output: html_notebook
---

```{r chunk_setup, include = FALSE, eval = TRUE}
knitr::opts_chunk$set(message = F, warning = F, comment = NA, 
                      cache.rebuild = F, cache = T, R.options = list(width = 120), 
                      fig.width = 8, fig.align = 'center')
```

## Effect of pharmacological actvation on outcome variables
### Outcome variables
- TEPs: 5 components (= amplitudes)
- MEPs: only 1 amplitude 
- RS-EEG: 3 related variables

### Factors
- MEDICATION: alprazolam X placebo
- TIME: baseline X post medication
- STIMULUS: CS X TS

### Group by
- Participant --> individual (inter-personal) variability --> participant RE
- Session --> intra-personal variability
- Stimulus --> only for TEPs, to cover for the fact that all TEP components are related


## Questions for Mme Bugli
- Shall I calculate with differences or should I make factor TIME: pre X post
- How do I add rMT as covariate?
  - rMT change --> to correct for changing of rMT
  - rMT --> to correct for variable intensity in stimulation = variable peripheral effects

## Statistical pipeline
Load packages
```{r basic_packages}
library(tidyverse)
library(lme4)
library(car)
library(ggplot2)
library(MASS)
library(dplyr)
library(ggpubr)
library(corrplot) 
library(visreg) 
library(rgl)
library(knitr)
library(scatterplot3d)
library(merTools)
library(readr)
```


Import data in R-compatible long format
```{r}
setwd('E:/UCL/O365G-NOCIONS - People/dsulcova/GABA-AD_results/Data')
data = read.csv('GABA_med_long.csv')
head(data, 10)
```

 
```{r}


```