---
title: 'GABA-AD: Interaction of SICI and pharmacological GABAAR activation'
output:
  html_notebook:
    theme:
      bootswatch: minty
    toc: yes
  html_document:
    df_print: paged
    toc: yes
chunk_output_type: inline
---

```{r chunk_setup, include = FALSE, eval = TRUE}
knitr::opts_chunk$set(echo = T, message = F, warning = F, comment = NA, autodep = F, eval = T, cache.rebuild = F, cache = T, R.options = list(width = 120), fig.width = 8, fig.align = 'center', dev.args = list(bg = 'transparent'), dev = 'svglite')
```

## **STUDY DESCRIPTION**
### Aim and hypothesis
The project aims to identify specific changes in our primary outcome measures (TMS-evoked potentials, TEPs) following the pre-activation of local network of cortical interneurons using a paired-pulse TMS protocol. The conditioning time (CS) of a sub-threshold intensity (80%rMT) his used to depolarize the low-threshold inhibitory interneuron at the site of the stimulation, which leads to a decrease in cortical excitability and a weaker response to the subsequent testing time (TS). This phenomenon is called the Short-latency IntraCortical Inhibition (SICI) and was originally measured in motor evoked potentials.

However, we can measure brain’s reaction to TMS more directly with TEPs, and we expect to see changes that will correspond to the increased local GABAA neurotransmission. The size of this effect can be then used to assess the efficiency of the system specifically at the site of stimulation. 

### Experimental design
20 subjects participated in 2 experimental sessions. Each session the same measures were taken at the baseline and after medication with either alprazolam (an allosteric BDZ agonist of GABAARs) or placebo. 

The question we aim to answer with our data are as follow:
Does the global activation of GABAARs with alprazolam affect the extent of SICI?

### Outcome variables
- **SICI**: calculated as the change of the SICI of each component (simple subtraction) following the piared-pulse stimulation as compared to the simple TS --> this measure serves to evaluate the effect of medication on SICI

### Fixed effect
- **TIME** 
  0. baseline 
  1. post-medication
- **MEDICATION**  
  0. placebo = control condition
  1. amplrazolam
  
- Covariate: **rMT change** - continuous value

### Random effect
- random intercept for each **SUBJECT**

**<span style = "color: red;">Following pipeline treats each outcome measure separately, as an independent measure.</span>**

## **STATISTICAL ANALYSIS: BASELINE SICI**

```{r setup, include = FALSE, eval = TRUE}
# report 3 significant digits
options(digits = 4)    

# load packages
library(skimr)
library(moments)
library(rstatix)
library(tidyverse)
library(ggpubr)
library(corrplot) 
library(visreg) 
library(knitr)
library(readr)
library(nlme)
library(lme4)
library(lmerTest)
library(pbkrtest)
library(car)

# directories
folder_dir <- choose.dir(default = getwd(), caption = "Select folder for input/output")
figures_dir <- paste(folder_dir, '\\figures', sep = "")
if(file.exists(figures_dir)){
  setwd(folder_dir)
}else {
    dir.create(figures_dir)
  }

# color palette - red/blue/purple/green(light + dark)
col_palette = c('cadetblue1', 'turquoise4', 'tomato1', 'red3', 'mediumorchid1', 'darkmagenta', 'olivedrab3', 'olivedrab4')
```

### Import data
Choose a .csv file with either TEP or rsEEG data. 
```{r data}
# import the data
input_file <- choose.files(default = folder_dir, caption = "Select the dataset")
data_all <- as_tibble(read.csv(input_file))
names(data_all)
summary(data_all)
```

### Variable of interest
```{r variable}
# variable characteristics
target_peak <- 'N17'

# subset data
data <- data_all %>% 
  filter(peak == target_peak) %>% 
  mutate(subject = factor(subject)) %>% 
  mutate(medication = factor(medication, levels=c('placebo', 'alprazolam'), ordered = TRUE)) %>% 
  mutate(time = factor(time, levels=c('pre', 'post'), ordered = TRUE)) %>% 
  select(subject, medication, time, SICI, rMT)
  # select(subject, medication, time, SICI_MEP, rMT) %>%
  # mutate(SICI = SICI_MEP)
str(data)
```

### Descriptive statistics 
The distribution is described and visualized for both levels of the factor 'medication' separately.
The normality is checked both visually (histogram, QQ-plot) and using the Shapiro-Wilk normality test. 
```{r desctiptives}
# ---- DISTRIBUTION ----
# get summary per medication
data %>% 
  select(c(medication, time, SICI)) %>% 
  group_by(medication, time) %>% 
  skim()

# calculate mean values
data_mean <- data %>% 
  group_by(medication, time) %>% 
  summarise(mean = mean(SICI), sd = sd(SICI), sem = sd(SICI)/sqrt(20), median = median(SICI))
data_mean %>% 
  knitr::kable()

# plot the distribution
data %>% 
  ggplot(aes(x = SICI, y = ..count.., fill = medication)) +
  geom_histogram(binwidth = 1, col = 'white', alpha = 0.5) +
  geom_density(alpha = 0, size = 1) +
  facet_grid(medication ~ time) + 
  geom_vline(data = data_mean, aes(xintercept = median), linetype = 'dashed', size = 1) + 
  geom_vline(data = data_mean, aes(xintercept = mean), size = 1) + 
  ggtitle('original data') +
  ylab('count') +
  xlab('SICI') + 
  theme(text = element_text(size = 15), legend.position = 'none') + 
  scale_fill_manual(values = col_palette[c(2, 4)]) 

# skewness and kurtosis
data_shape <- data %>% 
  group_by(medication, time) %>% 
  summarise(skewness = skewness(SICI), kurtosis = kurtosis(SICI))
data_shape %>% 
  knitr::kable()

# ---- NORMALITY TEST ----
# QQ-plot
data %>% 
  ggplot(aes(sample = SICI, col = medication)) + 
  geom_qq(size = 2.5) + 
  facet_grid(medication ~ time) + 
  scale_color_manual(values = col_palette[c(2, 4)]) 

# Shapiro-Wilk test
data %>% 
  group_by(medication, time) %>% 
  shapiro_test(SICI)

```

### Data transformation
Transformations are applied only when needed - in the case of a non-normal distribution.
The code below tests two possible transformations - square root and log transform with the base 2 - and their effect on the final distribution.
```{r transformation}
# constants --> onset to positive values
c <- -min(data$SICI) + 1
data_trans <- data %>% 
  # filter(!(subject %in% remove_subj)) %>%
  mutate(SICI = SICI + c)

# reflect if negatively skewed
# data_trans <- data_trans %>%
#   group_by(time, medication) %>%
#   mutate(SICI = max(data_trans$SICI) - SICI + 1)

# ---- SQUARE ROOT ----
# create new variable
data_trans <- data_trans %>% 
  mutate(SICI_sqrt = sqrt(SICI))

# plot by mediation
data_trans %>% 
  ggplot(aes(x = SICI_sqrt, y = ..count.., fill = medication)) +
  geom_histogram(binwidth = 1, col = 'white', alpha = 0.5) +
  geom_density(alpha = 0, size = 1) +
  facet_grid(medication ~ time) + 
  ggtitle('square-rooted data') +
  ylab('count') +
  xlab('SICI') + 
  theme(text = element_text(size = 15), legend.position = 'none') + 
  scale_fill_manual(values = col_palette[c(2, 4)]) 

# Shapiro-Wilk test
data_trans %>% 
  group_by(medication, time) %>% 
  shapiro_test(SICI_sqrt)

# ---- LOG TRANSFORM ----
# create new variable
data_trans <- data_trans %>% 
  mutate(SICI_log2 = log2(SICI))

# plot by mediation
data_trans %>% 
  ggplot(aes(x = SICI_log2, y = ..count.., fill = medication)) +
  geom_histogram(binwidth = 1, col = 'white', alpha = 0.5) +
  geom_density(alpha = 0, size = 1) +
  facet_grid(medication ~ time) + 
  ggtitle('log-transformed data') +
  ylab('count') +
  xlab('SICI') + 
  theme(text = element_text(size = 15), legend.position = 'none') + 
  scale_fill_manual(values = col_palette[c(2, 4)]) 

# Shapiro-Wilk test
data_trans %>% 
  group_by(medication, time) %>% 
  shapiro_test(SICI_log2)
```

### Outlier rejection
If a serious outlier is identified, the subject might be removed from the dataset. This step is an option if transformations don't prove helpful.
```{r outlier identification}
# outlier threshold 
out_threshold <- data %>% 
  group_by(medication, time) %>% 
  summarize(a = quantile(SICI, 0.75) - 1.5*IQR(SICI), b = quantile(SICI, 0.75) + 1.5*IQR(SICI))
out_threshold %>% 
  knitr::kable()

# idenntify outliers
outliers <- boxplot.stats(data %>% group_by(time, medication) %>% pull(SICI))$out
outliers_index <- which(data %>% group_by(time, medication) %>% pull(SICI) %in% c(outliers))
data[outliers_index,]

# boxplot
data %>% 
  ggplot(aes(time, SICI, col = medication)) + 
  geom_jitter(width = 0.2, size = 2.5, show.legend = FALSE) + 
  scale_color_manual(values = col_palette[c(2, 4)]) + 
  geom_boxplot(col = 'black', alpha = 0, width = 0.5) + 
  ylab('SICI') +
  xlab('medication') + 
  theme(text = element_text(size = 15))
```

```{r outlier rejection}
# choose which outliers to remove
# remove_subj <- data$subject[outliers_index[c(1)]]
remove_subj <- 2

# ---- DISTRIBUTION ----
# get summary per medication
data_filt <- data %>% 
  filter(!(subject %in% remove_subj)) %>% 
  select(c(medication, time, SICI)) %>% 
  group_by(medication, time)

# calculate mean values
data_mean <- data_filt %>% 
  group_by(medication, time) %>% 
  summarise(mean = mean(SICI), sd = sd(SICI), sem = sd(SICI)/sqrt(19), median = median(SICI))
data_mean %>% 
  knitr::kable()

# plot the distribution
data_filt %>% 
  ggplot(aes(x = SICI, y = ..count.., fill = medication)) +
  geom_histogram(binwidth = 1, col = 'white', alpha = 0.5) +
  geom_density(alpha = 0, size = 1) +
  facet_grid(medication ~ time) + 
  geom_vline(data = data_mean, aes(xintercept = median), linetype = 'dashed', size = 1) + 
  geom_vline(data = data_mean, aes(xintercept = mean), size = 1) + 
  ggtitle('original data') +
  ylab('count') +
  xlab('SICI') + 
  theme(text = element_text(size = 15), legend.position = 'none') + 
  scale_fill_manual(values = col_palette[c(2, 4)]) 

# skewness and kurtosis
data_shape <- data_filt %>% 
  group_by(medication, time) %>% 
  summarise(skewness = skewness(SICI), kurtosis = kurtosis(SICI))
data_shape %>% 
  knitr::kable()

# ---- NORMALITY TEST ----
# QQ-plot
data_filt %>% 
  ggplot(aes(sample = SICI, col = medication)) + 
  geom_qq(size = 2.5) + 
  facet_grid(medication ~ time) + 
  scale_color_manual(values = col_palette[c(2, 4)]) 

# Shapiro-Wilk test
data_filt %>% 
  group_by(medication, time) %>% 
  shapiro_test(SICI)
```

### Build models and compare
Based on the previous exploration, choose appropriate adjustment to the data.
```{r final data}
# choose transformation if necessary
trans <- readline(prompt = "Use transformation: ")
if(trans == 'sqrt'){
  data <- data %>%
    mutate(SICI_model = data_trans$SICI_sqrt)
}
if(trans == 'log'){
  data <- data %>%
  mutate(SICI_model = data_trans$SICI_log2)
}
if(trans == 'none'){
  data <- data %>%
  mutate(SICI_model = SICI)
}

# remove outliers if necessary
out <- askYesNo('Do you want to filter out ouliers?')
if(out){
  # data <- filter(data, !subject %in% remove_subj)
  data <- filter(data, !subject %in% 2)
}
```

Model n. 1: simple linear regression without random effects
```{r  regression}
model_lm <- lm(SICI_model ~ time + medication + time*medication,
  data = data
)
summary(model_lm)
confint(model_lm)
```

Model n. 2: empty model with random intercept
```{r random intercept}
model_ri <- lmer(
  SICI_model ~ 1 + (1|subject),
  data = data
)
summary(model_ri)
confint(model_ri)
```

Model n. 3: predictor variables + random intercept
```{r predictors}
model_rip <- lmer(
  SICI_model ~ time + medication + time*medication + rMT + (1|subject),
  data = data
)
summary(model_rip)
confint(model_rip)
```

```{r compare models}
anova(model_ri, model_rip)
```


Describe the full model using the nlme package in order to extract the p-value of the predictor. 
```{r final model}
# choose the model to retain
model <- lme(
  SICI_model ~ time + medication + time*medication + rMT,
  random = ~ 1|subject,
  data = data,
  method = 'ML'
)

# fixed effects
broom.mixed::tidy(model, effects = 'fixed') %>% 
  knitr::kable()

# fixed effects
broom.mixed::tidy(model, effects = 'ran_pars') %>% 
  knitr::kable()

# residuals
plot(model)
qqnorm(model)
```

Estimate p values using the Kenward-Roger approximation
```{r anova}
anova(model_rip, ddf="Kenward-Roger") %>% 
  knitr::kable()
```